---
title: "DH_BIO720_Assignment4"
output:
keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##The world of data visualization

Two basic types of data visualizations (R uses both)
1. Exploratory: visualizations help us understand data
-keep as much detail as possible 

2. Explanatory: visualizatons help us share our understanding with other people
-requires you to make choices ( ex. what features do you want to highlight) and edit (ex. remove distracting features)

There is 4 graphic systems in R 
1. Base Graphics: Easiest to learn
2. Grid Graphics: powerful set of modules for building other tools. Offers more power and flexibilty than base. Steeper learning curve and more effort required. 
3. Lattice Graphics:general purpose system based on grid graphics. Allows you to make conditional scatter plots. 
4. ggplot2: "the grammar of graphics". you can add details and modify. 

##Creating an exploratory plot array


```{r}
Deepsea2<-read.table("/Users/DarcyHenderson/Desktop/DeepSea2.txt", header = TRUE)
str(Deepsea2)
plot(Deepsea2)

```

- this plot is exploratory as it plots the entire DeepSea2 data frame and allows you too see what is important information and information that is not necessary to plot  

##Creating an explanatory scatterplot 

```{r}

plot(x =Deepsea2$Station, y = Deepsea2$BottomDepth, xlab = " Station", ylab = " Bottom Depth")

```

- this plot is explanatory as it plots information we have deemed important from the exploratory plot. It is a much more specific view and helps to show information we feel is important to view/share. 

##The plot() function is generic

- plot is generic which means results depend on oject applied to it 
- above produced scatter but what happens if factor variable is passed through


```{r}
microarraysample<-read.table("/Users/DarcyHenderson/Desktop/microarray_sample.txt", header = TRUE)
plot(microarraysample$group)

```

-plot function produces bar plot 

##A preview of some more and less useful techniques

#sunflowerplots
-are good when numerical data points are repeated as scatter plots do not show this 
- 1 petal = each repeated data point 

#Bloxplot
-used to show how the distrubution of a numerical varible changes over the discrete levels of a categorical variable
-also useful in showing relationship between a numerical y variable that assumes many different values and a numerical x variable that assumes only a few values 

#Mosaicplots
-designed to show relationships between two categorical variables 
-but can also be used to show relationships between two numerical variables with few values 

#High level 
-example is plot 

#Low level
 -points(): adds points 
 -lines(): adds lines -> usually curved 
 -text(): adds labels 
 
 par(): sets many graphic parameters 
  mfrow() : allows you to set up multiple plot arrays 
  

##Adding details to a plot using point shapes, color, and reference lines

pch argument: can add different shapes 
col argument: can add different colours 
points(): adds points to existing scatterplot 
aline(): adds reference line 

```{r}

source("http://www.openintro.org/stat/data/present.R") 


str(present)
plot( x= present$year, y= present$boys, pch = 17, col = "red")
points(x= present$year, y= present$girls, pch = 16, col = "blue")
abline(a = 0, b = 1, lty = 2)

```

##Creating multiple plot arrays

-when using plot() some options can or must be specified globally 
-other options can only be specified locally 

par(mfrow()):allows you to plot multiple graphs on a single pan

example: par(mfrow = c(1, 2)) creates a plot array with 1 row and 2 columns

```{r}

par(mfrow = c(1,2))

plot(x = Deepsea2$SampleID, y = Deepsea2$Latitude)
title("Original representation")

plot(x = Deepsea2$SampleID, y = Deepsea2$Latitude, log = "xy")
title("Log-log plot")

```

##Avoid pie charts

-If you are thinking about using a pie chart, it is probably a bad idea 

```{r}


par(mfrow = c(1,2))
tbl <- sort(table(Deepsea2$Latitude),
            decreasing = TRUE)
pie(tbl)
title("Pie chart")

barplot(tbl, las = 2, cex.names = 0.5)
title("Bar chart")


```


##The hist() and truehist() functions

```{r}
library(MASS)

par(mfrow = c(1,2))
hist(Deepsea2$BottomDepth, main = "hist() plot")

truehist(Deepsea2$BottomDepth, main= "truehist() plot")

```


##Density plots as smoothed histograms

gives a better estimate of the density function for a random variable

```{r}
cars_df <- read.csv("/Users/DarcyHenderson/Desktop/cars.csv")

index16 <- which(cars_df$Year == "70")

HP_70 <- cars_df$Horsepower[index16]

truehist(HP_70)
lines(density(HP_70))

```

##Using the qqPlot() function to see many details in data

```{r}
library("car")

qqPlot(HP_70)

qqPlot(Deepsea2$Time)

```

-notice differences. 
-in plot with HP_70 the Gaussian distribution appears to be reasonable
-in plot with Deepsea2 it is not reasonale

##The sunflowerplot() function for repeated numerical data

```{r}

par(mfrow = c(1,2))
plot(Horsepower ~ Year, data = cars_df)
 title("Standard scatterplot")

 sunflowerplot(Horsepower ~ Year, data = cars_df)
 title("Sunflower plot")
 
```

-reinforces that scatterplots aren't the best choice when you have repeated data points, as this isn't shown clearly
-sunflowerplots are able to show repeated data points 

##Useful options for the boxplot() function

-varwidth allows for variable-width boxplots that show the different sizes of the data subsets.
-log allows for log-transformed y-values.
-las allows for more readable axis labels.

```{r}

boxplot( Horsepower ~ Year, 
  data = cars_df, 
  varwidth = T,
  log = "y",
  las = 1)

title("Horsepower vs. Year")

```


-box plots that are wider are much larger subsets
-in this examples they are around the same size so subsets are around same size 

##Using the mosaicplot() function

-scatterplot between categorical variables
- useful in observing and understanding relationship between numerical variables, that take only a few different values

```{r}

mosaicplot(carb ~ cyl, 
  data = mtcars,
  color = T)

```

##Using the bagplot() function

A single box plot gives a graphical representation of the range of variation in a numerical variable, based on five numbers:

-The minimum and maximum values
-The median (or "middle") value
-Two intermediate values called the lower and upper quartiles

library(aplpack)

par(mfrow = c(1,2))
boxplot(microarraysample$g1, microarraysample$g2)

bagplot(microarraysample$g1, microarraysample$g2,
  cex = 1.20)

abline(0,1, lty = 2)

-problem loading aplpack package 
-cex makes point size larger

##Plotting correlation matrices with the corrplot() function

Correlation matrices are useful for obtaining a preliminary view of the relationships between multiple numerical variables

-ellipses that are thin and elongated indicate a large correlation value between the indicated variables
-ellipses that are nearly circular indicate correlations near zero

```{r}

library(corrplot)

str(microarraysample)

col_numeric <- sapply(microarraysample, is.numeric)

numericalVars <- microarraysample[,col_numeric]

str(numericalVars)

corrMat <- cor(numericalVars)
corrMat
corrplot(corrMat, method = "ellipse")

```

-g4 is correlated with g3

##Building and plotting rpart() models
- decision trees represent a popular form of predictive model

```{r}

library(rpart)

tree_model <- rpart(medv ~ ., data = Boston)

plot(tree_model)

text(tree_model, cex = .7)

```

##Introduction to the par() function

parameters may be set by a call in the form par(name = value) where name is the name of the parameter to be set and value is the value to be assigned to this parameter.

```{r}

plot_pars <- par()
names(plot_pars)
length(plot_pars)

```

-simply shows what these graphics parameters are

##Exploring the type option

-mfrow specifies the numbers of rows and columns in an array of plot
-values are two-element numerical vectors

values for how plot is made: 

"p" for "points"
"l" for "lines"
"o" for "overlaid" (i.e., lines overlaid with points)
"s" for "steps"

```{r}

tb_rates<-read.csv("/Users/DarcyHenderson/Desktop/tb_rates.csv", header = TRUE)
par(mfrow = c(2,2))

plot(tb_rates$X2000, type = "p")
title("points")

plot(tb_rates$X2000, type = "l")
title("lines")


plot(tb_rates$X2000, type = "o")
title("overlaid")

plot(tb_rates$X2000, type = "s")
title("steps")

```

##The surprising utility of the type "n" option

type = "n"
-purposefor when plotting data from multiple sources on a common set of axes
- can compute ranges for the x- and y-axes so that all data points will appear on the plot

```{r}

max_hp <- max(Cars93$Horsepower, mtcars$hp)
max_mpg <- max(Cars93$MPG.city, Cars93$MPG.highway, mtcars$mpg)
plot(Cars93$Horsepower, Cars93$MPG.city,
     type = "n", 
     xlim = c(0, max_hp),
     ylim = c(0, max_mpg), 
     xlab = "Horsepower",
     ylab = "Miles per gallon")

points(mtcars$hp, mtcars$mpg, pch = 1)
points(Cars93$Horsepower, Cars93$MPG.city, pch = 15)
points(Cars93$Horsepower, Cars93$MPG.highway, pch = 2)

```

##The lines() function and line types

```{r}

x <- seq(0, 10, length = 200)

gauss1 <- dnorm(x, mean = 2, sd = 0.2)

gauss2 <- dnorm(x, mean = 4, sd = 0.5)

plot(x, gauss1, 
  type = "l",
  ylab = "Gaussian probability density")

lines(x, gauss2, lty = 2, lwd = 3)
```

-shows what this bell curve looks like for exactly Gaussian data
-shows how the lines() function can be used to add lines to plot

 -line types are set by the lty
 -lty = 1 specifying solid lines
 -lty = 2 specifying dashed lines
 -lty = 3 specifying dotted lines
 -lwd argument specifies the width

##The points() function and point types

```{r}
plot(mtcars$hp, mtcars$mpg,
  type='n',
  xlab = "Horsepower",
  ylab = "Gas mileage")

points(mtcars$hp, mtcars$mpg, pch = mtcars$cyl)

plot(mtcars$hp, mtcars$mpg, 
  type='n',
  xlab = "Horsepower",
  ylab = "Gas mileage")

points(mtcars$hp, mtcars$mpg, pch = as.character(mtcars$cyl))

```

-first plot specifies the point shapes using numerical values of the pch argument defined by the cyl variable in the mtcars data frame
-The second plot shows that pch can also be specified as a vector of single characters, causing each point to be drawn as the corresponding character

##Adding trend lines from linear regression models

abline() adds a straight line to an existing plot
-line is specified by an intercept parameter a and a slope parameter b
-but can also specify these parameters through a linear regression model that determines them from data

1.generate a scatterplot of y versus x 
2.then fit a linear model that predicts y from x
3.call abline() to add this best fit line to the plot

```{r}

linear_model <- lm(Gas ~ Temp, data = whiteside)
plot(whiteside$Temp, whiteside$Gas)
abline(linear_model, lty = 2)

```
-dashed line shows general trend of data

##Using the text() function to label plot features

text() is used to add labels 
-x specifies the value for the x variable
-y specifies the value for the y variable
-label specifies the label for the x-y value pair

```{r}

plot(Cars93$Horsepower, Cars93$MPG.city, pch = 15)

head(Cars93)
index3 <- which(Cars93$Cylinders == 3)
index3

text(x = Cars93$Horsepower[index3],
     y = Cars93$MPG.city[index3], 
     labels = Cars93$Make[index3], 
     adj = 0)

```

##Adjusting text position, size, and font

-adj argument can change horizontal text placement -> value between 0-1 
-making adj (-) causes text to start to right of specified position
-making adj (+) causes text to start to left of specified postion 

-cex scales the default text size
-for example cex = 0.5 increases text size by 50%

-font is used to specify text font 
-font = 1 is default 
-font = 2 is bold
-font = 3 is italic 
-font = 4 is bold and italic 

```{r}

plot(Cars93$Horsepower, Cars93$MPG.city, pch = 1)

index3 <- which(Cars93$Cylinders == 3)

points(Cars93$Horsepower[index3], Cars93$MPG.city[index3], pch = 16)

text(
  Cars93$Horsepower[index3], 
  Cars93$MPG.city[index3], 
  Cars93$Make[index3], 
  adj = -0.2,
  cex = 1.2,
  font = 4)

```

-label is much more clear with adjustments 

##Rotating text with the srt argument

-srt argument allows you to rotate the text
srt = 90 causes the text to be rotated 90 degrees counter-clockwise

```{r}
plot(whiteside$Temp, whiteside$Gas, pch=17)

indexB <- which(whiteside$Insul == "Before")

indexA <- which(whiteside$Insul == "After")

text(x = whiteside$Temp[indexB], y = whiteside$Gas[indexB],
     labels = "Before", col = 'blue', srt = 30, cex = .8)

text(x = whiteside$Temp[indexA], y = whiteside$Gas[indexA],
     labels = "After", col = 'red', srt = -20, cex = .8)
```

##Using the legend() function

legends() adds explanatory text to a plot

```{r}
plot(whiteside$Temp, whiteside$Gas,
     type = "n", 
     xlab = "Outside temperature",
     ylab = "Heating gas consumption")

indexB <- which(whiteside$Insul == "Before")

indexA <- which(whiteside$Insul == "After")

points(whiteside$Temp[indexB], whiteside$Gas[indexB], pch = 17)

points(whiteside$Temp[indexA], whiteside$Gas[indexA], pch = 1)

legend("topright", pch = c(17, 1), legend = c("Before", "After"))

```

##Adding custom axes with the axis() function

When you want to add custom axes you need to:
1. suppress default axes
2. call function axis()

The side argument tells the function which axis to create
-value of 1 adds an axis below the plot
-2 adds an axis on the left
-3 puts it across the top
-4 puts it on the right side. 

 -at is a vector that defines points where tick-marks will be drawn on the axis

-labels is a vector that defines labels at each of these tick-marks

```{r}

boxplot(sugars ~ shelf, data = UScereal, axes = F)

axis(side = 2)

axis(side = 1, at = c(1,2,3))

axis(side = 3, at = c(1,2,3),
     labels = c("floor","middle","top"))

```

##Using the supsmu() function to add smooth trend curves

sometimes scatterplots show trends that are not linear, so what do you do?

-supsmu() allows for a curved line to be added 

-The bass argument controls the degree of smoothness in the resulting trend curve

```{r}

plot(Cars93$Horsepower, Cars93$MPG.city)

trend1 <- supsmu(Cars93$Horsepower, Cars93$MPG.city)

lines(trend1)

trend2 <- supsmu(Cars93$Horsepower, Cars93$MPG.city, bass = 10)

lines(trend2, lty = 3, lwd = 2)
```

##Too much is too much

when you use the plot() function to a data frame with many columns, it will generate an enormous array of scatterplots that are small and not useful 

```{r}

ncol(Deepsea2)^2
plot(Deepsea2)

```

##Deciding how many scatterplots is too many

-multiple plot arrays can be extremely helpful in illustrating: 
1. difference between datasets
2. different views of same dataset
3.similarities between datasets 
4.related views of the same dataset 

matplot() generates a plot with several scatterplots on the same set of axe

```{r}
keep_vars <- c("calories", "protein", "fat",
               "fibre", "carbo", "sugars")

df <- UScereal[, keep_vars]
str(df)
par(mfrow = c(2,2))

matplot(
  UScereal$calories, 
  UScereal[,c('protein', 'fat')], 
  xlab = "calories",
  ylab = "")

title("Two scatterplots")

matplot(
  UScereal$calories, 
  UScereal[,c('protein', 'fat', 'fibre')], 
  xlab = "calories",
  ylab = "")

title("Three scatterplots")

matplot(
  UScereal$calories, 
  UScereal[,c('protein', 'fat', 'fibre', 'carbo')], 
  xlab = "calories",
  ylab = "")

title("Four scatterplots")

matplot(
  UScereal$calories, 
  UScereal[,c('protein', 'fat', 'fibre', 'carbo','sugars')], 
  xlab = "calories",
  ylab = "")

title("Five scatterplots")

```

- as you can see in this adding to many scatterplots is not useful

##How many words is too many?

-scatterplot arrays can become not useful if they become too complex 

wordcloud() is a function that is used to create wordclouds 
-present words in varying sizes depending on their frequency
-called with character vector of words and second numerical vector giving the number of times each word appears 

-scale argument is a two-component numerical vector giving the relative size of the largest word in the display and that of the smallest word
-wordcloud only includes those words that occur at least min.freq times in the collection and the default value for this argument is 3

```{r}

library(wordcloud)

mfr_table <- table(Cars93$Manufacturer)

wordcloud(
  words = names(mfr_table), 
  freq = as.numeric(mfr_table), 
  scale = c(2, 0.25))

wordcloud(
  words = names(mfr_table), 
  freq = as.numeric(mfr_table), 
  scale = c(2, 0.25),
  min.freq = 1)

model_table <- table(Cars93$Model)

wordcloud(
  words = names(model_table), 
  freq = as.numeric(model_table), 
  scale = c(.75, .25), 
  min.freq = 1)
```

##The Anscombe quartet

-datasets can appear very similar when comparing summary statistics 

##The utility of common scaling and individual titles

-you can see the differences in these datasets if you plot scatterplots with the same x and y ranges

##Using multiple plots to give multiple views of a dataset

-multiple plot arrays also let you see multiple related views of the same dataset 
-gives a deeper understanding of a data variable

##Constructing and displaying layout matrices

-layout() function creates plot array 
-0 represents empty space and other numbers represent the plot number 
ex. a 1 in the layout matrix reders to the visualization that was first called 

-create matrix with three rows and two columns
-1st row is empty plot to the left and plot 1 to the right 
-2nd row is plot 2 to the left and empty plot to right
-3rd row is empty plot to ledt and plot 3 to right 

```{r}

layoutMatrix <- matrix(
  c(
    0, 1,
    2, 0,
    0, 3
  ), 
  byrow = T, 
  nrow = 3
)

layout(layoutMatrix)

layout.show(n = 3)
```

##Creating a triangular array of plots

-can create three differnt views of data.frame 
-sometimes it is not possible to construct a plot array using the mfrow parameter(as it is used for rectangular arrays only)

```{r}

layout(layoutMatrix)

indexB <- which(whiteside$Insul == "Before")
indexA <- which(whiteside$Insul == "After")

plot(whiteside$Temp[indexB], whiteside$Gas[indexB],
     ylim = c(0,8))
title("Before data only")


plot(whiteside$Temp, whiteside$Gas,
     ylim = c(0,8))
title("Complete dataset")

plot(whiteside$Temp[indexA], whiteside$Gas[indexA],
     ylim = c(0,8))
title("After data only")

```

##Creating arrays with different sized plots
-layout() can be used to create plot arrays with different sized component plots 

```{r}

row1 <- c(1,0,0)
row2 <- c(0,2,2)
layoutVector <- c(row1,row2, row2)

layoutMatrix <- matrix(layoutVector, byrow = T, nrow = 3)

layout(layoutMatrix)

plot(Boston$rad, Boston$zn)

sunflowerplot(Boston$rad, Boston$zn)
```

##Some plot functions also return useful information

-barplot() returns a numerical vector with the center position of each bar in the plot
-this vector is normally returned invisibly but you can observe it 
-this vector is useful when you want to overlay text on the bars of a horixontal barplot 

```{r}
tbl <- table(Cars93$Cylinders)

mids <- barplot(tbl, 
  horiz = T,
  col = "transparent",
  names.arg = "")

text(20, mids, names(tbl))
text(35, mids, as.numeric(tbl))
```

##Using the symbols() function to display relations between more than two variables

-symbols() allows you to show influence of other variables
-x and y define scatterplot
-circles argument creates bubbleplot 
-each data point is represented by a circle whose radius depends on the third variable

```{r}
symbols(Cars93$Horsepower, Cars93$MPG.city,
        circles = sqrt(Cars93$Price))
        
symbols(Cars93$Horsepower, Cars93$MPG.city,
        circles = sqrt(Cars93$Price),
        inches = 0.1)

```

-notice differences in size of circles

##Saving plot results as files
-png() function sets the name of the png file to be generated and sets up a special environment that captures all graphical output until you exit environment with the dev.off() command

##Iliinsky and Steele's 12 recommended colors
IScolors <- c("red", "green", "yellow", "blue",
              "black", "white", "pink", "cyan",
              "gray", "orange", "brown", "purple")

##Using color to enhance a bubbleplot
-colour can be used in bubbleplots to illustrate differences 
```{r}

IScolors <- c("red", "green", "yellow", "blue",
              "black", "white", "pink", "cyan",
              "gray", "orange", "brown", "purple")
cylinderLevel <- as.numeric(Cars93$Cylinders)

symbols(
  Cars93$Horsepower, Cars93$MPG.city, 
  circles = as.numeric(Cars93$Cylinders), 
  inches = 0.2, 
  bg = IScolors[cylinderLevel])
```

##Using color to enhance stacked barplots
-Stacked bar plots can be made using the barplot()
-each bar in the plot is partitioned into segments characterizing portions of the data characterized by the bar

```{r}
tbl <- table(Cars93$Cylinders, Cars93$Origin)
barplot(tbl)
barplot(tbl, col = IScolors)
```

-notice difference when using colour 

##The tabplot package and grid graphics
-tablot is a package that allows you to use grid graphics 
-tableplot() allows you to visualize data distributions and relationships between variables in large data sets 
-> constructs side by side horizonatal barplots, for each variable 
-> works best for viewing up to 10 

##A lattice graphics example
-built on grid graphics 
-lattice graphics and base graphics are similar functions but can produce very different results
-lattice graphics supports conditional plots -> show relationships between variables within different groups

```{r}
library(lattice)
calories_vs_sugars_by_shelf <- calories ~ sugars | shelf
xyplot(calories_vs_sugars_by_shelf, data = UScereal)
```

##A ggplot2 graphics example
-grid graphics 
-based on the grammar of graphics  -> approach to building and changing graphical displays. Start with basics and add to it through use of modifiers 

```{r}
library(ggplot2)

basePlot <- ggplot(Cars93, aes(x = Horsepower, y = MPG.city))

basePlot + 
  geom_point()
```
- This is basic plot 
-can now add to it 

```{r}
 basePlot + 
  geom_point(colour = IScolors[Cars93$Cylinders])
```
-adds colour 

```{r}
  basePlot + 
  geom_point(colour = IScolors[Cars93$Cylinders], 
             size = as.numeric(Cars93$Cylinders))
```
-makes point size vary 

##Why Base R? 
-flexible
-good for exploratory analysis 
-easy to learn 

##Why Use Grid Graphics?
-greater control over low level graphical details 
-more flexible than base graphics 
-comes at cost of steep learning curve 


